---
title: "Project 1"
author: "Dario Molina, Evert Rodriguez"
date: "March 28, 2018"
output: html_document
---
<!--
Intro: explain what you will do in a sentence or two
Data: Explain in a few sentence where the data comes from, what it is about, etc.
Initial data exploration
Data cleaning and preprocessing
Data exploration and visualization
Conclusions.  One paragraph about what you learned, issues you faced, etc.
-->

###Introduction
In this project, we will look into Car Collision Report data, and analyze in detail what this data represents.  We will make graphs, to visually analyze and find patterns, make predictions, and come to a better understanding on how we can reduce the amount of car collisions.

The data is obtained from dataMontgomery(https://data.montgomerycountymd.gov/Public-Safety/Crash-Reporting-Drivers-Data/mmzv-x632/data), a website that keeps records from the County of Maryland.

```{r global_options, include=FALSE}
knitr::opts_chunk$set(prompt=TRUE, comment="", echo=FALSE)
```


```{r }
#File locaiton on Dario's Macine
#dat = read.csv("/Users/dariomolina/DataScience/Crash-Report-Project/Crash_Reporting_-_Drivers_Data.csv",stringsAsFactors=FALSE)
#Address on Evert's Machine

dat = read.csv("/Users/evert/Dropbox/School/Assignments/Crash_Reporting_-_Drivers_Data.csv",stringsAsFactors=FALSE)

```

###Data cleaning and preprocessing

Changing empty blanks to 0 in Off.Road.Description, Related.Non.Motorist, Non.Motorist.Substance.Abuse columns
```{r}
x = c("Off.Road.Description", "Related.Non.Motorist", "Non.Motorist.Substance.Abuse")

#assigning value 0 to "" in x vector
for( i in x)
{
  dat[[i]][dat[[i]] == ""] = 0
}

```

Removing rest of empty blanks and "N/A" in data set to N/A
```{r}
#assignig rest values to NA
for(i in names(dat))
{
  dat[[i]][dat[[i]] == "" | dat[[i]] == "N/A"] = NA
}

```

###Most Common Type of Crashes
```{r}
par(mar=c(3,14,2,2))
x = table(dat$Collision.Type)
barplot(x[x > 5000], horiz=TRUE, las=1, main="Most common types of car crash", col="firebrick")
```

###Amount of Crashes per Month
```{r}
#amount of crashes per month
x =table(substr(dat$Crash.Date.Time,1,2))
names(x) = c("Jan.","Feb.","Mar.","Apr.","May","Jun.","Jul.","Aug.","Sep.","Oct.","Nov.","Dec.")
barplot(x, las = 1, col="firebrick")
```

###Most Common Types of Crashes Under Different Types of Weather
```{r}
z = table(dat$Weather,dat$Collision.Type)
y= z[,c("STRAIGHT MOVEMENT ANGLE","SINGLE VEHICLE","SAME DIRECTION RIGHT TURN","SAME DIRECTION LEFT TURN","HEAD ON LEFT TURN","OTHER")]
y = y[c("CLEAR","CLOUDY","RAINING"),]

yCol = c("STRGT-MV-AGL", "ONE-CAR","SM-DIR-R-TR","SM-DIR-L-TR","HD-ON-L-TR","OTHER")
for(i in 1:length(yCol))
{
  colnames(y)[i] = yCol[i]
}
y = y[,c("STRGT-MV-AGL","ONE-CAR","HD-ON-L-TR","OTHER")]
par(mar=c(1,.5,2,.3))
plot(y, col="firebrick")
```


###Different Types of Crashes, Most Common Report Type
```{r}
#Whattypes of crash types caused the most reports by category

  #remove coluns from crash type 
y = table(dat$ACRS.Report.Type, dat$Collision.Type)
x= c("STRAIGHT MOVEMENT ANGLE","SINGLE VEHICLE","SAME DIR REAR END","OTHER","HEAD ON LEFT TURN")
y = y[,x]

yCol = c("STRGT-MV-AGL", "ONE-CAR","SM-DIR-R-ED","OTHER","HD-ON-L-TR")
for(i in 1:length(yCol))
{
  colnames(y)[i] = yCol[i]
}
par(mar=c(0.5,2,1,2))
plot(y, las=1,col="firebrick")
```

###Most Common Types of Crash Type based on Surface Condition
```{r}
#crashes by surface condition
yCol = c("HD-ON-L-TR","OTHER","SM-DIR-R-ED","SM-DIR-SDWE","ONE-CAR","STRGT-MV-AGL")

x = table(dat$Surface.Condition,dat$Collision.Type)


cols = c("HEAD ON LEFT TURN","OTHER","SAME DIR REAR END","SAME DIRECTION SIDESWIPE","SINGLE VEHICLE","STRAIGHT MOVEMENT ANGLE")
rows = c("DRY","WET","SNOW")
z = x[rows,cols]

for(i in 1:length(yCol))
{
  colnames(z)[i] = yCol[i]
}

par(mar=c(1,.01,2,.01))
plot(z, las=1,col="firebrick")
```

###More Data Cleaning and Pre-Prosessing
Creating a new dataframe containing incidents in which a bicyclist was involved and the driver was at fault and NOT distracted. (194 data points)

```{r }
dat_bi_daf = dat[dat$Related.Non.Motorist == "BICYCLIST" & dat$Driver.At.Fault == "Yes",]

dat.sub1 <- subset(dat_bi_daf, Driver.Distracted.By == "LOOKED BUT DID NOT SEE")
dat.sub2 <- subset(dat_bi_daf, Driver.Distracted.By == "NOT DISTRACTED")

dat.sub <- rbind(dat.sub1,dat.sub2)


```

###Bicyle-Involved Accidents Over the Year
To start, we were intrested in accidents involving bicycles.  We decided to only look at the incidents in which the driver was at fault and not distracted because these situations are the ones where we can hopefully correct.  This gave us almost 200 data points to analyze.  We did some inital plots to try and find some pattern to these accidents.This first graph shows the total number of accidents that occured for each month.

```{r }
par(mar=c(2,9,2,2))
x =table(substr(dat.sub$Crash.Date.Time,1,2))
names(x) = c("Jan.","Feb.","Mar.","Apr.","May","Jun.","Jul.","Aug.","Sep.","Oct.","Nov.","Dec.")
barplot(x, las = 1, col="blue4", main="Num. of Accidents Per Month")
```

Not too surprisingly, the amount of crashes increases throughout the year as the weather is more pleasurable and dips down during the winter season.

###Frequency of Bicycle-Involved Crashes Over the Day
After the frist graph, we wondered if the number crashes would also increase during times of heavy traffic during a day, such as when people are commuting to/from work/school.

```{r}
#formating the Crash.Date.Time Column
dat.sub$Crash.Date.Time <- as.POSIXlt(as.character(dat.sub$Crash.Date.Time), format = "%m/%d/%Y %I:%M:%S %p")
par(mar=c(4,2.5,2,2))
y = table(substr(dat.sub$Crash.Date.Time,12,13))
names(y) = c("12am","1am","5am","6am","7am","8am","9am","10am","11am","12pm","1pm","2pm","3pm","4pm","5pm","6pm","7pm","8pm","9pm","10pm","11pm")
barplot(y,las=1,col = "blue4",main="Total Numb. of Accidents Per Each Hour", xlab = "Hour")
```

The graph confirmed our inital suspisions by showing that the number of crashes spiked during common comute times.

###Location of Bicycle-Involved Crashes

Next, we decided to see if we could gain any insights in regards to where these accidents happen so we plotted them onto a map of the county.

```{r results='hide', message=FALSE, warning=FALSE}
library(ggmap)
map = get_map(c(-77.08025,39.07991),maptype = "road",zoom=11)
p = ggmap(map)
p = p + geom_point(data=dat.sub, aes(x=Longitude,y=Latitude),color="blue4",size=3)
print(p)

```

Although not too obvious from the small graph, it seemed to indicate that a significant amount of the incidents occured at or near an intersection.

###Most Common Type of Vehicle Movement in Bicycle-Involved Accidents
Based on the results of the previous map, we decided to graph the movement of the vehicle.

```{r}
par(mar=c(2,13,2,2))
barplot(head(sort(table(dat.sub$Vehicle.Movement),decreasing = TRUE)),col="blue4",las=1,horiz = TRUE,main="Vehicle Movement")
```

From the plot, we learned that about 1/3 of the incidents involved the vehicle making some sort of turn.

###Location of Incidents by Vehicle Movement

Finally, we decided to generalize the vehicle movement category into four categories in order to simplify maping the points. The turning sub-category indicates incidents in which the vehicle was making any type of turn (left,right,U), the accelerating sub-category shows incidents where the vehicle was either changing speed (slowing dow/speeding up) or changing lanes (including merging into traffic).' 

```{r results='hide', message=FALSE, warning=FALSE}

turning=c("MAKING LEFT TURN", "MAKING RIGHT TURN","MAKING U TURN","RIGHT TURN ON RED")

acc = c("ACCELERATING","ENTERING TRAFFIC LANE","STARTING FROM PARKED","STARTING FROM PARKED","STARTING FROM LANE", "SLOWING OR STOPPING","CHANGING LANES")

other = c("BACKING","SKIDDING", "PASSING" , "MOVING CONSTANT SPEED")

stat = c("PARKED","STOPPED IN TRAFFIC LANE")

dat.sub$Vehicle.Movement[dat.sub$Vehicle.Movement %in% turning] <- "TURNING"
dat.sub$Vehicle.Movement[dat.sub$Vehicle.Movement %in% acc] <- "ACCELERATING"
dat.sub$Vehicle.Movement[dat.sub$Vehicle.Movement %in% other] <- "OTHER"
dat.sub$Vehicle.Movement[dat.sub$Vehicle.Movement %in% stat] <- "STATIONARY"


map = get_map(c(-77.08025,39.07991),maptype = "road",zoom=11)
p = ggmap(map)
p = p + geom_point(data=dat.sub, aes(x=Longitude,y=Latitude, color = Vehicle.Movement),size=3)

print(p)

```